---
title: "Chapter 2"
description: |
  Reproducing Chapter 2 of Bayesian Methods for Hackers in R + Stan
author:
  - name: J. Duncan
date: 01-08-2021
output:
  distill::distill_article:
    self_contained: false
---
## 2.2 Modeling Approaches

### 2.2.10 Example: *Challenger* Space Shuttle Disaster

```{r}
library(tidyverse)
library(cmdstanr)
library(posterior)
library(bayesplot)
library(tidybayes)
library(latex2exp)
library(modelr)

url <- "https://raw.githubusercontent.com/CamDavidsonPilon/Probabilistic-Programming-and-Bayesian-Methods-for-Hackers/master/Chapter2_MorePyMC/data/challenger_data.csv"

data <- read_csv(url, col_types = "cdi") %>% 
  janitor::clean_names() %>% 
  mutate(
    date = if_else(
      nchar(date) >= 10,
      parse_date(date, format = "%m/%d/%Y"),
      parse_date(date, format = "%m/%d/%y")
      )
    ) %>% 
  drop_na()
```

Let's visualize this data to make sure all is well:

```{r fig.height=3}
theme_set(theme_tidybayes())

data %>% 
  ggplot(aes(x=temperature, y=damage_incident)) + 
  geom_point(alpha = 0.5) +
  labs(
    title = "Defects of the Space Shuttle O-Rings vs temperature",
    x = "Outside temperautre (Farenheit)",
    y = "Damage Incident"
  ) +
  scale_y_continuous(breaks = scales::pretty_breaks(n = 1)) +
  xlim(50, 85)
```

https://jrnold.github.io/bayesian_notes/binomial-models.html

```{r fig.height=3}
map_df(
  list(-1, -3, 5),
  ~ tibble(
    x = seq(-4, 4, length.out = 101),
    y = plogis(x * .x),
    Parameter = paste0("Beta: ", .x)
    )
  ) %>%
  ggplot(aes(x = x, y = y, colour = Parameter)) +
  geom_line() 
```

### 2.2.11 The Normal Distribution

```{r}
data_list <- data %>% 
  rename(damage = damage_incident) %>% 
  compose_data()
```

```{r}
mod <- cmdstanr::cmdstan_model("models/ch2-mod.stan")
mod$print()

fit <- mod$sample(
  data = data_list,
  seed = 123,
  chains = 4,
  parallel_chains = 4,
  refresh = 1000,
  iter_sampling = 3000
)

fit$summary()
```

https://mpopov.com/blog/2020/09/07/pivoting-posteriors/

```{r}
tidy_draws.CmdStanMCMC <- function(model, ...) {
  return(as_draws_df(model$draws()))
}
```

```{r}
fit %>% 
  gather_draws(alpha, beta) %>% 
  ggplot(aes(x = .value)) +
  geom_histogram() +
  facet_wrap(~.variable, scales = "free", ncol = 1) +
  labs(
    title = TeX("Posterior Distributions of $\\alpha$ and $\\beta$"),
    y = NULL
  )
```

Using comments from [discourse](https://discourse.mc-stan.org/t/best-way-to-do-prediction-on-new-data-r-rstan-stanfit/1772):

> Extract posterior parameter samples into R and do prediction in R with new-data-for-prediction.

```{r}
post_df <- spread_draws(fit, alpha, beta)
t <- seq(min(data$temperature) - 5, max(data$temperature) + 5, length.out = 50)

pred_df <- tibble(temperature = t) %>% 
  mutate(nest(post_df)) %>% # nest all posterior parameters samples
  mutate(
    data = map2(
      data, 
      temperature, 
      ~ mutate(.x, pred = plogis(alpha + beta * .y)) # predict
    )
  ) %>% 
  unnest(data)
```

Now for the visualizations. First let's look at realizations and then will summarize all of the realizations:

```{r fig.height=4}
pred_df %>% 
  group_by(temperature) %>% 
  sample_draws(100) %>% 
  ggplot(aes(x = temperature)) +
  geom_line(aes(y = pred, group = .draw), alpha = 0.3) + 
  geom_point(data = data, aes(y = damage_incident), alpha = 0.3) +
  labs(
    title = "Posterior probability estimates given temp; realizations",
    y = "probability estimate"
  )
```

Now for all of the realizations:

```{r fig.height=4}
pred_df %>% 
  group_by(temperature) %>% 
  median_hdci(pred, .width = c(0.95, 0.8)) %>% 
  ggplot(aes(x = temperature)) +
  geom_lineribbon(aes(y = pred, ymin = .lower, ymax = .upper), color = "#08519C") +
  geom_point(data = data, aes(y = damage_incident), alpha = 0.5) +
  scale_fill_brewer() +
  labs(
    title = "Posterior probability estimates given temp",
    y = "probability estimate"
  )
```

### 2.2.12 What about the day of the *Challenger* disaster?

> On the day of the Challenger disaster, the outside temperature was `31` degrees Fahrenheit. What is the posterior distribution of a defect occurring, given this temperature? The distribution is plotted below. It looks almost guaranteed that the Challenger was going to be subject to defective O-rings.

```{r}
prob_31 <- post_df %>% 
  mutate(pred = plogis(alpha + beta * 31))

prob_31 %>% 
  ggplot(aes(x = pred)) +
  stat_dots(quantiles = 25) +
  labs(
    title = TeX("Posterior distribution of probability of defect, given $t = 31$"),
    x = "probability of defect occurring in O-ring",
    y = NULL
  )
```

## 2.3 Is Our Model Appropriate?
### 2.3.1 Separation Plots

Where *expected* number of defects for the Bernoulli random variable is given as:

$$
E[S] = \sum_{i=0}^N E[X_i] = \sum_{i=0}^N p_i
$$
Preparing our data for the separation plot:

```{r}
sep_data <- fit %>% 
  spread_draws(yrep[day]) %>% 
  mean_hdci(yrep) %>% 
  left_join(mutate(data, day = row_number())) %>% 
  arrange(yrep) %>% 
  mutate(idx = row_number())

exp_def <- sep_data %>% 
  summarize(exp_def = sum(yrep)) %>% 
  pull()
```

Now let's review the separation plot:

```{r fig.height=4}
sep_data %>% 
  ggplot(aes(x=idx)) +
  geom_col(aes(y=damage_incident), alpha = 0.3) +
  geom_step(aes(y = yrep)) +
  geom_vline(xintercept = 23 - exp_def, linetype = "dotted") +
  labs(
    title = "Separation Plot",
    subtitle = paste0("with ", round(exp_def, 1), " expected defects"),
    y = NULL,
    x = NULL
  )
```

> The snaking-line is the sorted probabilities, ~~blue~~ gray bars denote defects, and empty space denote non-defects. As the probability rises, we see more and more defects occur. On the right hand side, the plot suggests that as the posterior probability is large (line close to 1), then more defects are realized. This is good behaviour. Ideally, all the ~~blue~~ gray bars should be close to the right-hand side, and deviations from this reflect missed predictions.
